A general framework for learning planning action models

### 2018-06-12 doc/aij18/compilation.tex
updates added to aij18 paper

\subsection{Tuning the compilation with background knowledge}
A distinctive feature of Inductive Logic Programming (ILP) is that ILP can leverage {\em background knowledge} to learn logic programs from data~\cite{muggleton1994inductive}. Inspired by ILP, we show that our approach for the learning of \strips\ action models can also leverage {\em background knowledge} in this case to optimize the performance of the $P_{\Lambda}$ compilation.

\subsubsection{Static predicates}
A {\em static predicate} $p \in \Psi$ is a predicate that does not appear in the effects of any action~\cite{fox:TIM:JAIR1998}. Therefore, one can get rid of the mechanism for programming these predicates in the effects of any action schema while keeping the compilation complete. Given a static predicate $p$:
\begin{itemize}
\item Fluents $del_f(\xi)$ and $add_f(\xi)$, such that $f\in F_v$ is an instantiation of the static predicate $p$ in the set of {\em variable objects} $\Omega_v$, can be discarded for every $\xi\in\Xi$.
\item Actions $\mathsf{programEff_{f,\xi}}$ (s.t. $f\in F_v$ is an instantiation of $p$ in $\Omega_v$) can also be discarded for every $\xi\in\Xi$.
\end{itemize}

@@

\subsubsection{State constraints}
The notion of {\em state-constraint} is very general and has been used in different areas of AI and for different purposes.  If we restrict ourselves to planning, {\em state-constraints} are abstractions for compactly specifying sets of states. For instance, {\em state-constraints} in planning allow to specify the set of states where a given action is applicable, the set of states where a given {\em derived predicate} holds or the set of states that are considered goal states.

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$.

The formula $\phi_{I,A}^*$ represents the {\em strongest invariant} and exactly characterizes the set of all states reachable from $I$ with the actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em strongest invariant} for {\em blocksworld}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard as hard as testing plan existence.

\begin{figure}[hbt!]
  \begin{footnotesize}
    \begin{center}
$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
    \end{center}
\end{footnotesize}
 \caption{\small An example of the strongest invariant for the {\em blocksworld} domain.}
\label{fig:strongest-invariant}
\end{figure}

A {\em mutex} (mutually exclusive) is a state invariant that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\phi_1=\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a mutex because $block_A$ can only be on top of a single block.

A {\em domain invariant} is an instance-independent invariant, i.e. holds for any possible initial state and set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants)~\cite{rintanen:schematicInvariants:AAAI2017}. For instance, $\phi_2=\forall x:\ (\neg handempty\vee \neg holding(x))$, is a {\em domain mutex} for the {\em blocksworld} because the robot hand is never empty and holding a block at the same time.

An interesting contribution of our compilation is that the validation of an action model can also be done with {\em state constraints}. Given $\Phi$, a set of either state or trajectory constraints, our validate actions can be adapted to check also that the learned model satisfy a constraint $\phi\in\Phi$:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&\phi\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq |\Phi|}\cup \{mode_{val}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
\end{align*}
\end{small}

This redefinition of validate actions applies also to {\em trajectory constraints} because LTL formulae can be represented using classical action preconditions and goals by encoding the Non-deterministic B\"{u}chi Automaton (NBA), that is equivalent to the corresponding LTL formula, as part of the classical planning tasks~\cite{baier2006planning}.

Because of the combinatorial nature of the search for a solution plan, the sooner unpromising nodes are pruned from the search the more efficient the computation of a solution plan. Constraints can be used to confine earlier the set of possible \strips\ action models and reduce then the learning hypothesis space. With regard to our compilation, {\em domain mutex} are useful to reduce the amount of applicable actions for programming a precondition or an effect for a given action schema. For example given the {\em domain mutex} $\phi=(\neg f_1\vee \neg f_2)$ such that $f_1\in F_v(\xi)$ and $f_2\in F_v(\xi)$, we can redefine the corresponding programming actions for {\bf removing} the {\em precondition} $f_1\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$ as:

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{f_1,\xi}})=&\{\neg del_{f_1}(\xi),\neg add_{f_1}(\xi), mode_{prog}, pre_{f_1}(\xi), pre_{f_2}(\xi)\},\\
\cond(\mathsf{programPre_{f_1,\xi}})=&\{\emptyset\}\rhd\{\neg pre_{f_1}(\xi)\}.
\end{align*}
\end{small}

>>>>

\subsection{Optimizing the compilation with background knowledge}
A distinctive feature of Inductive Logic Programming (ILP) is that ILP can leverage {\em background knowledge} to learn logic programs from data~\cite{muggleton1994inductive}. Inspired by ILP, we show that our approach for the learning of \strips\ action models can also leverage {\em background knowledge} in this case to optimize the performance of the $P_{\Lambda}$ compilation.

%\subsubsection{Static predicates}
A {\em static predicate} $p \in \Psi$ is a predicate that does not appear in the effects of any action~\cite{fox:TIM:JAIR1998}. Therefore, one can get rid of the mechanism for programming these predicates in the effects of any action schema while keeping the compilation complete. Given a static predicate $p$:
\begin{itemize}
\item Fluents $del_f(\xi)$ and $add_f(\xi)$, such that $f\in F_v$ is an instantiation of the static predicate $p$ in the set of {\em variable objects} $\Omega_v$, can be discarded for every $\xi\in\Xi$.
\item Actions $\mathsf{programEff_{f,\xi}}$ (s.t. $f\in F_v$ is an instantiation of $p$ in $\Omega_v$) can also be discarded for every $\xi\in\Xi$.
\end{itemize}

@@

%\subsubsection{State constraints}
%The notion of {\em state-constraint} is very general and has been used in different areas of AI and for different purposes.  If we restrict ourselves to planning, {\em state-constraints} are abstractions for compactly specifying sets of states. For instance, {\em state-constraints} in planning allow to specify the set of states where a given action is applicable, the set of states where a given {\em derived predicate} holds or the set of states that are considered goal states.

%{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$.

%The formula $\phi_{I,A}^*$ represents the {\em strongest invariant} and exactly characterizes the set of all states reachable from $I$ with the actions in $A$. For instance Figure~\ref{fig:strongest-invariant} shows five clauses that define the {\em strongest invariant} for {\em blocksworld}. There are infinitely many strongest invariants, but they are all logically equivalent, and computing the strongest invariant is PSPACE-hard as hard as testing plan existence.

%\begin{figure}[hbt!]
%  \begin{footnotesize}
%    \begin{center}
%$\forall x_1,x_2\ ontable(x_1)\leftrightarrow\neg on(x_1,x_2)$.\\
%$\forall x_1,x_2\ clear(x_1)\leftrightarrow\neg on(x_2,x_1)$.\\
%$\forall x_1,x_2,x_3\ \neg on(x_1,x_2)\vee\neg on(x_1,x_3)\ such\ that\ x_2\neq x_3$.\\
%$\forall x_1,x_2,x_3\ \neg on(x_2,x_1)\vee\neg on(x_3,x_1)\ such\ that\ x_2\neq x_3$.\\
%$\forall x_1,\ldots,x_n\ \neg(on(x_1,x_2)\wedge on(x_2,x_3)\wedge\ldots\wedge on(x_{n-1},x_n)\wedge on(x_n,x_1)).$
%    \end{center}
%\end{footnotesize}
% \caption{\small An example of the strongest invariant for the {\em blocksworld} domain.}
%\label{fig:strongest-invariant}
%\end{figure}

%A {\em mutex} (mutually exclusive) is a state invariant that takes the form of a binary clause and indicates a pair of different properties that cannot be simultaneously true~\cite{kautz:mutex:IJCAI1999}. For instance in a three-block {\em blocksworld}, $\phi_1=\neg on(block_A,block_B)\vee \neg on(block_A,block_C)$ is a mutex because $block_A$ can only be on top of a single block.

%A {\em domain invariant} is an instance-independent invariant, i.e. holds for any possible initial state and set of objects. Therefore, if a given state $s$ holds $s\nvDash \phi$ such that $\phi$ is a {\em domain invariant}, it means that $s$ is not a valid state. Domain invariants are often compactly defined as {\em lifted invariants} (also called schematic invariants)~\cite{rintanen:schematicInvariants:AAAI2017}. For instance, $\phi_2=\forall x:\ (\neg handempty\vee \neg holding(x))$, is a {\em domain mutex} for the {\em blocksworld} because the robot hand is never empty and holding a block at the same time.

%An interesting contribution of our compilation is that the validation of an action model can also be done with {\em state constraints}. Given $\Phi$, a set of either state or trajectory constraints, our validate actions can be adapted to check also that the learned model satisfy a constraint $\phi\in\Phi$:
%\begin{small}
%\begin{align*}
%\hspace*{7pt}\pre(\mathsf{validate_{i}})=&\phi\cup\{test_j\}_{j\in 1\leq j<i}\cup\{\neg test_j\}_{j\in i\leq j\leq |\Phi|}\cup \{mode_{val}\},\\
%\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i,\neg mode_{val}\}.
%\end{align*}
%\end{small}

%This redefinition of validate actions applies also to {\em trajectory constraints} because LTL formulae can be represented using classical action preconditions and goals by encoding the Non-deterministic B\"{u}chi Automaton (NBA), that is equivalent to the corresponding LTL formula, as part of the classical planning tasks~\cite{baier2006planning}.

%Because of the combinatorial nature of the search for a solution plan, the sooner unpromising nodes are pruned from the search the more efficient the computation of a solution plan. Constraints can be used to confine earlier the set of possible \strips\ action models and reduce then the learning hypothesis space. With regard to our compilation, {\em domain mutex} are useful to reduce the amount of applicable actions for programming a precondition or an effect for a given action schema. For example given the {\em domain mutex} $\phi=(\neg f_1\vee \neg f_2)$ such that $f_1\in F_v(\xi)$ and $f_2\in F_v(\xi)$, we can redefine the corresponding programming actions for {\bf removing} the {\em precondition} $f_1\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$ as:

%\begin{small}
%\begin{align*}
%\hspace*{7pt}\pre(\mathsf{programPre_{f_1,\xi}})=&\{\neg del_{f_1}(\xi),\neg add_{f_1}(\xi), mode_{prog}, pre_{f_1}(\xi), pre_{f_2}(\xi)\},\\
%\cond(\mathsf{programPre_{f_1,\xi}})=&\{\emptyset\}\rhd\{\neg pre_{f_1}(\xi)\}.
%\end{align*}
%\end{small}

### 2018-06-12 doc/aij18/evaluation.tex
updates added to aij18 paper

If the actual model is available, the quality of a learned action model is quantifiable using the {\em precision} and {\em recall} metrics. {\em Precision} and {\em recall} are two syntactic metrics frequently used in {\em pattern recognition}, {\em information retrieval} and {\em binary classification} and are more informative that simply counting the number of errors in the learned model or computing the {\em symmetric difference} between the learned and the reference model~\cite{davis2006relationship}. Intuitively, precision gives a notion of {\em soundness} while recall gives a notion of the {\em completeness} of the learned models:
\begin{itemize}
\item $Precision=\frac{tp}{tp+fp}$, where $tp$ is the number of {\em true positives} (predicates that correctly appear in the action model) and $fp$ is the number of {\em false positives} (predicates of the learned model that should not appear).
\item $Recall=\frac{tp}{tp+fn}$, where $fn$ is the number of {\em false negatives} (predicates that should appear in the learned model but are missing).
\end{itemize}

The roles of two {\em comparable} action schemes, or the roles of two action parameters that belong to the same type, can be swapped by a $\Lambda$ learning task when the observed input data, given in $\mathcal{T}$, is insufficient. For instance, when $\mathcal{T}$ does not contain any intermediate action the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa, or the parameters of the {\small\tt stack} operator can be swapped. Pure syntax-based evaluation metrics (like {\em precision} and {\em recall}) can report low scores for learned models that are actually good but correspond to {\em reformulations} of the actual model; i.e. a learned model semantically equivalent but syntactically different to the reference model.

Here we introduce a novel evaluation approach that is robust to role changes of this particular kind. The intuition of the approach is to {\em semantically} assess how well a \strips\ action model $\mathcal{M}$ explains given observations of plan executions according to the amount of {\em edition} required by $\mathcal{M}$ to induce that observations. This semantic evaluation approach is again flexible to various amount and kind of available input knowledge.

@@

Note that the number of deletions indicate the number of {\em false positives} while the number of insertions represents the number of {\em false negatives}. As explained, these numbers can be normalized with the amount of {\em true positives} to compute the {\em precision} and {\em recall} of a learned model with respect to a reference model. Normally evaluating a given learned domain with respect to the actual generative model is not possible because the actual model is not available. As the ARMS system shows, the error of a learned action model can also be estimated with respect to a set of observations of plan executions~\cite{yang2007learning}. With this regard, we define now an edit distance to asses the quality of a learned action model with respect to a plan trace $\mathcal{T}=\tup{s_0,a,_1,s_1,\ldots,a_n,s_{n}}$ obtained watching the execution of a plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applied in $s_{i-1}$ and generates the successor state $s_i$.

@@

The {\em observation edit distance} is mappable into a likelihood with the following expression $1-\frac{\delta(\mathcal{M},\mathcal{T})}{\delta(\mathcal{M},*)}$. Further, following the previous correspondances ({\em false positives} $\equiv$ {\em deletions} and {\em false negatives} $\equiv$ {\em insertions}) we can redefine a {\em semantic} version of the  {\em precision} and {\em recall} metrics of a learned model with respect to a partially observed plan trace $\mathcal{T}=\tup{s_0,a,_1,s_1,\ldots,a_n,s_{n}}$. In this case, for each action schema $\xi\in\mathcal{M}$, the corresponding number of {\em true positives} is given by $tp(\xi)=|pre(\xi)|+|del(\xi)|+|add(\xi)|-fp(\xi)$.

>>>>

If the actual GTM model is available then the quality of a learned action model is quantifiable using the ML metrics, {\em precision} and {\em recall}. {\em Precision} and {\em recall} are two syntactic metrics frequently used in {\em pattern recognition}, {\em information retrieval} and {\em binary classification} and are more informative than counting the number of errors between the learned and the reference model~\cite{davis2006relationship}. Intuitively, precision gives a notion of {\em soundness} while recall gives a notion of the {\em completeness} of the learned models:
\begin{itemize}
\item $Precision=\frac{tp}{tp+fp}$, where $tp$ is the number of {\em true positives} (in our case, predicates that correctly appear in the action model) and $fp$ is the number of {\em false positives} (predicates of the learned model that should not appear).
\item $Recall=\frac{tp}{tp+fn}$, where $fn$ is the number of {\em false negatives} (predicates that should appear in the learned model but are missing).
\end{itemize}

In a $\Lambda$ learning task, the roles of two {\em comparable} action schemes (or the roles of two action parameters with the same type) can be swapped. These role swaps tipically happen when the observed input data, given in $\mathcal{T}$, is scarce. For instance when $\mathcal{T}$ does not contain any intermediate action, the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa. Further, the roles of parameters of the {\small\tt stack} (or the {\small\tt unstack}) operator could be swapped and these learned models would still be sematically correct with respect to the given input observations.

Pure syntax-based evaluation metrics (like {\em precision} and {\em recall}) can report low scores for learned models that are actually {\em sound} and {\em complete} but correspond to {\em reformulations} of the actual model; i.e. a learned model semantically equivalent but syntactically different to the reference model. Here we introduce a novel evaluation approach that is robust to role changes of this particular kind. The intuition of the approach is to {\em semantically} assess how well a \strips\ action model $\mathcal{M}$ explains given observations of plan executions according to the amount of {\em edition} required by $\mathcal{M}$ to induce that observations. This semantic evaluation approach is again flexible to various amount and kind of available input knowledge.

@@

Note that the number of deletions indicate the number of {\em false positives} while the number of insertions corresponds to the number of {\em false negatives}. As explained, these numbers can be normalized with the amount of {\em true positives} to compute the {\em precision} and {\em recall} of a given learned model with respect to a reference model.

Normally evaluating a given learned domain with respect to the actual GTM model is not possible because the actual GTM model is unknown. As the ARMS system shows, the error of a learned action model can also be estimated with respect to a set of observations of plan executions that are assumed to be generated with that model~\cite{yang2007learning}. With this regard, we define now an edit distance to asses the quality of a learned action model with respect to a partially observed plan trace $\mathcal{T}$.

@@

Following the previous correspondances ({\em false positives} $\equiv$ {\em deletions} and {\em false negatives} $\equiv$ {\em insertions}) we can define a {\em semantic} version of the {\em precision} and {\em recall} metrics of a learned model. In this case {\em precision} and {\em recall} is not computed with respect to a GTM but with respect to a partially observed plan trace $\mathcal{T}=\tup{s_0,a,_1,s_1,\ldots,a_n,s_{n}}$. To define the {\em semantic} version of {\em precision} and {\em recall} the only term to reformalize is the number of {\em true positives} that, for each action schema $\xi\in\mathcal{M}$, is given by $tp(\xi)=|pre(\xi)|+|del(\xi)|+|add(\xi)|-fp(\xi)$.
